---
title: Session 8
subtitle: Free practical time!
date: "2021-07-01"
author:
  - Matt Denwood
theme: metropolis
aspectratio: 43
colortheme: seahorse
header-includes: 
  - \input{../rsc/preamble}
params:
  presentation: TRUE
output:
  beamer_presentation:
      pandoc_args: ["-t", "beamer"]
      slide_level: 2
  html_document: default
---

```{r rendering, eval=FALSE, include=FALSE}
# To render this as PDF (beamer) slides run:
rmarkdown::render('Session_8.Rmd', 'beamer_presentation', params=list(presentation=TRUE))
# And for html:
rmarkdown::render('Session_8.Rmd', 'html_document', params=list(presentation=FALSE))
```

```{r setup, include=FALSE}
library("tidyverse")
library("runjags")
set.seed(2021-06-22)

# Reduce the width of R code output for PDF only:
if(params$presentation) options(width=60)
knitr::opts_chunk$set(echo = TRUE)

# Reduce font size of R code output for Beamer:
if(params$presentation){
  knitr::knit_hooks$set(size = function(before, options, envir) {
    if(before){
      knitr::asis_output(paste0("\\", options$size))
    }else{
      knitr::asis_output("\\normalsize")
    }
  })
  knitr::opts_chunk$set(size = "scriptsize")
}

# Collapse successive chunks:
space_collapse <- function(x){ gsub("```\n*```r*\n*", "", x) }
# Reduce space between chunks:
space_reduce <- function(x){ gsub("```\n+```\n", "", x) }
knitr::knit_hooks$set(document = space_collapse)

# To collect temporary filenames:
cleanup <- character(0)
```

# Session 8:  Free practical time

## Before you go...

- Feedback on the course would be extremely welcome!
  - I will send an email later today asking for Stop/Start/Continue suggestions

- Remember to keep an eye on the COST action website:
  - http://harmony-net.eu
  - Physical training schools are being run in September and accepting sign-ups now!
  

## Suggested activities for the rest of the morning

1. Work on your own data where possible, or help others in groups

1. Go back and look at the optional exercises if you didn't get chance to complete them all

1. General discussion / Q&A session in plenum

. . .

We will be here until 12.30

. . .

Thanks for participating!


```{r include=FALSE}
unlink(cleanup)
```
